%
% bullet_ammo.tex
%
% Z Specification for the XBoing bullet/ammo subsystem
% Formal model of the authoritative ammo count, the ammo-belt UI display,
% and the end-of-level bullet bonus — built to expose a display/bonus
% desync bug in the unlimited-ammo (MAXAMMO) path.
%

\documentclass[a4paper,10pt,fleqn]{article}
\usepackage[margin=1in]{geometry}
\usepackage{fuzz}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}

\begin{document}

\title{XBoing Bullet/Ammo: A Z Specification}
\author{Formal Model of the Ammo Count, Belt Display, and Bullet Bonus}
\date{July 2026}
\hypersetup{
  pdftitle={XBoing Bullet/Ammo: A Z Specification},
  pdfauthor={xboing-c},
  pdfsubject={Z Specification},
  pdfcreator={fuzz/probcli}
}
\maketitle

\tableofcontents
\newpage

\section{Introduction}

This is a narrow model of the bullet/ammo subsystem, built to locate a
reported desync: the player fired bullets while the ammo belt showed
none, and the end-of-level bonus rewarded many bullets the player did
not appear to hold.

The model separates three quantities the code keeps distinct:

\begin{itemize}
\item \textbf{actualAmmo} --- the authoritative count owned by
      \texttt{gun\_system} (\texttt{gun\_system\_get\_ammo}).
\item \textbf{displayedAmmo} --- how many bullets the ammo-belt renderer
      draws in the HUD (\texttt{game\_render\_ammo\_belt}).
\item \textbf{bonusAmmo} --- the count the end-of-level bonus reads and
      rewards (\texttt{mode\_bonus\_enter} $\to$
      \texttt{bonus\_system}).
\end{itemize}

The MAXAMMO block sets \texttt{unlimited} and stores a sentinel count of
\texttt{GUN\_MAX\_AMMO + 1}.  In the \emph{as-coded} (buggy) renderer the
belt hid entirely when \texttt{unlimited} was set while the bonus read the
raw sentinel, so the property \texttt{displayedAmmo = bonusAmmo} --- what
you see equals what you are paid for --- was violated at a reachable
state.  This document models the \emph{fixed} rule
(\texttt{displayedAmmo = actualAmmo}) and carries that property as an
invariant the operations preserve; the as-coded rule and the reachable
violating state it produced are recorded in section~\ref{sec:bug}.

\section{Free Types}

\begin{zed}
ZBOOL ::= ztrue | zfalse
\end{zed}

\section{Global Constants}

\begin{axdef}
maxAmmo : \nat \\
ammoPerLevel : \nat \\
bulletsPerPickup : \nat \\
unlimitedSentinel : \nat
\where
maxAmmo = 20 \\
ammoPerLevel = 4 \\
bulletsPerPickup = 4 \\
unlimitedSentinel = maxAmmo + 1
\end{axdef}

\section{Gun State (fixed belt rule)}

\texttt{displayedAmmo} is a derived rendering of state, pinned by the
belt rule each frame.  The \emph{fixed} belt rule (matching the original
\texttt{ReDrawBulletsLeft}, \texttt{level.c:329}): draw the actual count,
with no \texttt{unlimited} check and no cap --- so the belt always
reflects the authoritative ammo.  \texttt{bonusAmmo} is a snapshot the
bonus screen takes; it is meaningful only once \texttt{inBonus} is set.
The as-coded (buggy) belt rule and the desync it caused are documented
in section~\ref{sec:bug}.

\begin{schema}{Gun}
actualAmmo : \nat \\
unlimited : ZBOOL \\
displayedAmmo : \nat \\
inBonus : ZBOOL \\
bonusAmmo : \nat
\where
actualAmmo \leq unlimitedSentinel \\
bonusAmmo \leq unlimitedSentinel \\
(unlimited = zfalse \implies actualAmmo \leq maxAmmo) \\
displayedAmmo = actualAmmo \\
(inBonus = ztrue \implies displayedAmmo = bonusAmmo)
\end{schema}

\begin{schema}{Init}
Gun'
\where
actualAmmo' = ammoPerLevel \\
unlimited' = zfalse \\
displayedAmmo' = ammoPerLevel \\
inBonus' = zfalse \\
bonusAmmo' = 0
\end{schema}

\section{Operations}

\subsection{Fire a bullet}

Guard: ammo available or unlimited.  Unlimited does not decrement
(\texttt{gun\_system\_use\_ammo} is a no-op when unlimited).

\begin{schema}{FireBullet}
\Delta Gun
\where
(actualAmmo > 0 \lor unlimited = ztrue) \\
\land (unlimited = ztrue \implies actualAmmo' = actualAmmo) \\
\land (unlimited = zfalse \implies actualAmmo' = actualAmmo - 1) \\
unlimited' = unlimited \\
inBonus' = inBonus \\
bonusAmmo' = bonusAmmo
\end{schema}

\subsection{Pick up a BULLET block (+4 ammo)}

Ignored when unlimited (matches \texttt{game\_callbacks.c}: the +4 loop
is skipped under unlimited).  Clamps at \texttt{maxAmmo}.

\begin{schema}{PickupBullets}
\Delta Gun
\where
unlimited = zfalse \\
\land (actualAmmo + bulletsPerPickup \leq maxAmmo \\
\quad~ \implies actualAmmo' = actualAmmo + bulletsPerPickup) \\
\land (actualAmmo + bulletsPerPickup > maxAmmo \\
\quad~ \implies actualAmmo' = maxAmmo) \\
unlimited' = unlimited \\
inBonus' = inBonus \\
bonusAmmo' = bonusAmmo
\end{schema}

\subsection{Pick up a MAXAMMO block (unlimited)}

Sets unlimited and stores the sentinel count.  Under the fixed belt rule
(\texttt{displayedAmmo = actualAmmo}) the belt then shows all
\texttt{unlimitedSentinel} (21) bullets, matching the original --- it does
not vanish.  (The as-coded rule that forced \texttt{displayedAmmo} to 0
here, and the desync it caused, are in section~\ref{sec:bug}.)

\begin{schema}{PickupMaxAmmo}
\Delta Gun
\where
actualAmmo' = unlimitedSentinel \\
unlimited' = ztrue \\
inBonus' = inBonus \\
bonusAmmo' = bonusAmmo
\end{schema}

\subsection{Enter the bonus screen}

The bonus reads the raw authoritative count --- no unlimited check, no
cap (\texttt{mode\_bonus\_enter}: \texttt{ammo = gun\_system\_get\_ammo};
\texttt{.bullet\_count = ammo}).

\begin{schema}{EnterBonus}
\Delta Gun
\where
inBonus = zfalse \\
inBonus' = ztrue \\
bonusAmmo' = actualAmmo \\
actualAmmo' = actualAmmo \\
unlimited' = unlimited
\end{schema}

\subsection{Advance to the next level}

Resets unlimited and grants the per-level ammo
(\texttt{game\_rules.c}: \texttt{set\_unlimited(0)};
\texttt{set\_ammo(ammoPerLevel)}).

\begin{schema}{NextLevel}
\Delta Gun
\where
actualAmmo' = ammoPerLevel \\
unlimited' = zfalse \\
inBonus' = zfalse \\
bonusAmmo' = 0
\end{schema}

\section{Consistency Properties}

\subsection{Display matches bonus}

What the player sees on the belt must equal what the bonus rewards.
This is the property the reported bug violated; under the fixed belt
rule it is an invariant of \texttt{Gun} and every operation preserves
it.

\begin{schema}{DisplayMatchesBonus}
Gun
\where
inBonus = ztrue \implies displayedAmmo = bonusAmmo
\end{schema}

\subsection{Bonus never exceeds what was shown}

The bonus must not reward more bullets than the belt displayed.  With
$displayedAmmo = actualAmmo$ and $bonusAmmo = actualAmmo$ at the bonus,
this holds with equality.

\begin{schema}{BonusNotAboveDisplay}
Gun
\where
inBonus = ztrue \implies bonusAmmo \leq displayedAmmo
\end{schema}

\subsection{Fireability is visible}

If the player can fire, the belt shows at least one bullet.  Under the
fix, unlimited keeps \texttt{actualAmmo} at the sentinel and
$displayedAmmo = actualAmmo > 0$, so the belt is never empty while
firing is allowed --- the direct negation of ``firing bullets when it
did not look like I had any''.

\begin{schema}{FireabilityVisible}
Gun
\where
(actualAmmo > 0 \lor unlimited = ztrue) \implies displayedAmmo > 0
\end{schema}

\section{The Bug (as-coded) and the Fix}
\label{sec:bug}

\textbf{As-coded belt rule (the bug).}  The renderer
(\texttt{game\_render\_ammo\_belt}) hid the belt entirely when unlimited
and capped otherwise:
\[
 displayedAmmo =
 \begin{cases}
   0 & \text{if } unlimited \\
   \min(actualAmmo, maxAmmo) & \text{otherwise.}
 \end{cases}
\]
The bonus (\texttt{mode\_bonus\_enter}) read the raw count
$bonusAmmo = actualAmmo$ with no unlimited check.  From \texttt{Init},
the sequence \texttt{PickupMaxAmmo} then \texttt{EnterBonus} reached
$actualAmmo = 21$, $unlimited$, $displayedAmmo = 0$, $bonusAmmo = 21$ ---
violating \texttt{DisplayMatchesBonus} ($0 \neq 21$),
\texttt{BonusNotAboveDisplay} ($21 \not\leq 0$), and (already after
\texttt{PickupMaxAmmo}) \texttt{FireabilityVisible}.  Model-checking the
as-coded belt rule, the property $inBonus \implies displayedAmmo =
bonusAmmo$ excluded 441 of 486 reachable states --- the desync region.

\textbf{The fix.}  Replace the belt rule with $displayedAmmo =
actualAmmo$ (no unlimited check, no cap), matching the original
\texttt{ReDrawBulletsLeft} (\texttt{level.c:329}).  This is the belt rule
in the \texttt{Gun} schema above; under it all three properties are
invariants preserved by every operation.

\section{System Invariants}

\begin{itemize}
\item Physical: $actualAmmo \leq unlimitedSentinel$; the fixed belt rule
      pins $displayedAmmo = actualAmmo$.
\item Restored (holds under the fix): $inBonus = ztrue \implies
      displayedAmmo = bonusAmmo$.
\item Root cause: three consumers interpreted \texttt{unlimited}
      differently.  The belt hid on unlimited (showed 0); firing treats
      unlimited as ``always allowed''; the bonus ignored unlimited and
      read the raw sentinel.  The fix makes the belt reflect the actual
      count the bonus rewards, restoring a single interpretation across
      all three.
\end{itemize}

\end{document}
